Interrogating the linear variable importance of local explanations of black-box models with animated linear projections

### 2021-12-10 _paper.tex
- restructure paper into classification/regression - new screencaptures and remove old ones - a lot of grammarly edits - abstract causing error

]{article}

@@

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models, one such approach is the use of local explanations. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation, and approximate data and this attribution space side-by-side with linked brushing. After identifying an observation of interest its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with a technique called the tour. This tour animates many projections over small changes in the projection basis. Doing so allows a user to visually explore the data space through the lens of this local explanation and interrogate its variable importance. The implementation of our framework is available as an \textbf{R} package \textbf{cheem} available at \href{https://github.com/nspyrison/cheem}{github.com/nspyrison/cheem}.
\end{abstract}

@@

Mathematically rigorous approaches to predictive modeling are attributed to the method of least squares, over two centuries ago by Legendre and Gauss in 1805 and 1809 respectively. In 1886 Francis Galton coined the term \emph{regression} to refer to continuous, quantitative predictions. While \emph{classification} refers to discrete predictions as introduced by Fisher in 1936.

Breiman and Shmueli (Breiman 2001; Shmueli 2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose such as hypothesis testing, while \emph{predictive} modeling is performed to predict new or future out-of-sample observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, the interpretable is a key feature for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use of a model has important implications for which methods are used and the development of those models.

Black-box models are becoming increasingly common, but not without their share of controversy and issues (O'neil 2016; Kodiyan 2019). Applications have been known to reflect common biases against sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018). Such issues occur when biases existent in the training data, the model picks up on this influence on the response variable which is then built into the model. Another issue is that of data drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions (Lazer et al. 2014; Salzberg 2014). Such issues highlight the need to make models fair, accountable, ethical, and transparent which has led to the movement of XAI (Adadi and Berrada 2018; Arrieta et al. 2020).

One branch of XAI is local explanations, which take a variable attribution approach to bring transparency to a model. Local explanations attempt to approximate linear variable importance at the location of one observation. There are many such local explanations, any of which is works with our approach (assuming model-explanation compatibility).

To illustrate our work we apply the model-agnostic explanation SHAP (Strumbelj and Kononenko 2010; Štrumbelj and Kononenko 2014). The exact details of SHAP are tangent to the ideas of this work, but suffice it to say that SHAP approximates variable importance by taking the median importance over permutations of the explanatory variables. To be exact we apply a variant that enjoys a lower computational complexity, known as tree SHAP (S. M. Lundberg, Erion, and Lee 2018).

In multivariate data visualization a \emph{tour} (Asimov 1985; Buja and Asimov 1986; S. Lee et al. 2021) is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of the structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

There are various types of tours, which are distinguished by the sequence of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020) this path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path (Y. D. Lee et al. 2013; Silva, Cook, and Lee 2021).

The approach purposed below is to use the manual tour as means to interrogate a local explanation; a means of evaluating if its variable importance is a good explanation for the model predictions. We provide a free and open-source R package \texttt{cheem} with an interactive application to facilitate analysis. By viewing approximations of data- and attribution-space side-by-side, with linked brushing an analyst can identify observations of interest whose explanations are then rendered at the initial projection basis and explored with a manual tour to further interpret the variable importance of the local explanation. We give case studies of toy and modern datasets for both classification and regression tasks.

The change in the projection basis might feel similar to counterfactual, what-if analysis, such as \emph{Ceteris paribus} (Biecek 2020). Latin for ``other things held constant'' or ``all else unchanged,'' is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual observed case by varying the basis, the angle of the data object. Another difference is that the basis is restricted on orthonormality. That is to say when the contribution of one variable decreases the contributions of others necessarily increases such that there is a full component in that direction.

The rest of this paper is organized as follows. The next section \protect\hyperlink{sec:explanations}{Local explanation statistics} covers the background of the local explanation SHAP and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into the animation and what it shows. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the layout of the application, how it facilitates analysis, and discusses the backend details of the package and preprocessing. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:discussion}{Discussion} of the insights we draw from classification and regression tasks.

@@

Consider a highly non-linear model. At face value it is hard to say which variable(s) are sensitive to crossing of classification boundary or identify which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

Figure 6 of Arrieta et al. (2020) gives a broad summarization of the taxonomy and literature of explanation techniques. This includes a large number of model-specific explanations such as deepLIFT, (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017) a popular recursive method for estimating importance in neural networks. There are a fewer number of model-agnostic explanations, of which LIME, (Ribeiro, Singh, and Guestrin 2016) SHAP, (S. Lundberg and Lee 2017) and their variants are popular.

These instance level explanations are used in a variety of ways depending on the data. In images saliency maps overlay or offset a heatmap indicating which pixels were important (Simonyan, Vedaldi, and Zisserman 2014). For instance, or the presence of snow may be highlighted when distinguishing if a picture of contains a wolf or husky. In text analysis, Word-level contextual sentiment analysis can be used to highlight sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they can be used to explain variable additive contributions from intercept to prediction (Ribeiro, Singh, and Guestrin 2016).

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation by taking the median importance of a subset of permutations in the explanatory variables. This idea stems from the field of game theory where Shapley devised a method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score (Shapley 1953).

To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season (Leone 2020). We have 5000 observations of nine skill measures (after aggregating variable with high correlation). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more important to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions as well, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry with respect to the variable ordering, so we opt to fix the order to that of panel b), namely, by decreasing the sum of the SHAP values.

@@

\caption{Illustration of the distribution of SHAP attributions and a breakdown plot. From FIFA 2020 data, a random forest model regresses wages from 9 skill attributes for a star offensive and defensive player. a) The players have very different wages. b) Shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The median of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the position of the player. c) Breakdown plots of the observations the explanation used to additively explain the difference between the intercept and prediction}\label{fig:shapdistrbd}
\end{figure}

In summary, this highlights how local explanations bring interpretability to a model at least in the vicinity of their observations. In this instance, we showed how two very different soccer players receive different profiles of variable importance to explain the prediction of their wages. In the following section, we will be using normalized explanations as the starting projection basis to interrogate the explanation further.

@@

Below we illustrate the two primary displays of the application: the global view and the tour view. Then we'll cover what we take away from the classification and regression tasks. Lastly, we discuss the preprocessing that needs to be done before display.

@@

What information do we glean from using this method on a classification task? Typically we select a misclassified observation in comparison with a correctly classified point that is nearby in data space. We start by seeing the data projected through the linear attribution, the combination that best justifies that prediction. By default, the manual tour varies the contribution of the variable with the largest difference between the primary and comparison observation. That is, we can test the sensitivity of each variable to structure identified by the local explanation, we are exploring the support of the explanation, evaluating the support or robustness of the prediction.

@@

The regression case is not as discrete a feature. Instead, we find it more meaningful to compare the observed value or residual with the projection. In this case the horizontal positions are the same, that is projection of the data through the attribution of the explanation and then varying as the radial tour changes the contributions of the selected variable. While the height of the points is fixed to the fixed either to the observed y and residuals for the middle and right panels respectively.

The global view is an important context for exploring the separability of the data- and the local explanation's attribution-spaces, and is crucial in the selection of explanation to further interrogate and explore the structural sensitivity of.

We show an approximation of these spaces with a projection through their first two principal components. The orientation of the variables is shown inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it provides a reasonable starting point for the real task at hand, the selection of observation and nearby comparison.

It is insightful to explore these two approximations against a visual of the model; prediction by observation is added. Linked brushing and preserved aesthetic features such as circling misclassified observations help link information from the different spaces together.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/global_view_toy_class}

@@

\caption{Global view screen capture; the approximations of the data and SHAP-spaces of Palmer penguins data. Orientation of the basis contributions is illustrated on a unit circle. Linked brushing allows observations brushed in one plot to be selected in others. This selection is also used in the proceeding view and their corresponding information is displayed in an interactive table. Hovering the cursor over an observation displays a tooltip with row number/name information. In the classification case, misclassified points are circled in red. This view provides an orientation to select a primary and comparison observation, key targets in the following tour.}\label{fig:globalview}
\end{figure}

This view offers dynamic interaction in several ways. A tooltip displays while the cursor hovers over a point displays the row number/name and classification information if appropriate. Linked brushing allows for the selection of points (by click and drag) where those points will be highlighted in both plots. The information corresponding to the selected points is populated on a sortable table and the data powering the proceeding tour will also subset the data to the current selection.

\hypertarget{cheem-tour}{%
\subsection{Cheem tour}\label{cheem-tour}}

The primary observation identified via the global view is foundational to the production of the cheem tour. Namely, the linear attribution of that variable is used as a 1D projection basis. This is the approximate contribution of the variables that this model uses to justify its prediction for the observation.

The SHAP values of the full dataset are shown as horizontal parallel coordinate plots at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of a bar is mapped to the variable's contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \ref{fig:cheemtourclass} we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/cheem_tour_toy_class}

@@

\caption{Cheem tour, classification case. A random forest model predicting the class of the of simulated data. A misclassified observation is be compared with a nearby correctly classified observation. The top shows density and rug marks of the current frame with the positions of the primary and comparison observations shown as dashed and dotted lines. The bottom shows two key features; the distribution of attribution space shown as parallel coordinate lines across the variables and the bars show the current contribution of the variable to the projection. The primary, misclassifed observation (dashed line), is plausibly in the middle of the its prediction, purple density, the story that the explanation trying to sell us. Yet, when we play the tour animating on the contribution of V1, the bottom dashed line is more regularly in the center the it observed cluster, green, By varying this contribution we gain information of how sensitive this variable is to the explanation; the variable importance which led to a misclassification.}\label{fig:cheemtourclass}
\end{figure}

Data visualization tours animate many linear projections over small changes to the basis. The manual tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing allows an analyst to test an individual variable's sensitivity to the structure identified in the frame. The default variable selected is the one with the largest discrepancy between the primary and comparison observation's attribution. In the following sections we elaborate on the takeaways we draw from applying this approach in classification and regression tasks respectively.

In the regression case, we compare an observation with a particularly extreme residual as compared with a neighboring point that is more accurately predicted. The global tour is primarily the same without the misclassified circles, but the tooltip now shows the observation's residual. We simulate data, five variables from the uniform distribution in the support {[}0, 5{]} and \(y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon\). We fit a random forest and extract the tree SHAP values from all observations. The first thing we notice from the global view is the shape difference between the data- and attribution-spaces, which are primarily circular are triangular respectively. We also see that the random forest fit the data quite well viewing the predictions across the observations. Though linked brushing we find that the attribution space is highly correlated with the \(y\) value. The highest values occupy one corner of the triangle, the lowest values cluster on the opposite side, and the middle values form a `)' shape filling out the triangle in attribution space. In the regression case, the tour view is a bit different, this is illustrated in figure \ref{fig:cheemtourreg}. The current frames basis and distribution are shown to the left. The horizontal positions of the two scatter plots are the resulting 1D projection of the basis while the heights are fixed to the observed y and the residuals.

{\centering \includegraphics[width=1\linewidth]{./figures/cheem_tour_toy_reg}

\caption{Cheem tour, regression case. A random forest model predicting the continuous \(y\) of the of simulated data where \(y = x1 + x2 + x1*x2 + (x3 + x4 + x5) / 10 + \epsilon\). The current frames basis and the distribution of all attributions is shown on the left. The horizontal positions for the scatterplots are resulting projection through the 1D basis while the vertical heights are fixed to the observed y and residuals respectively.}\label{fig:cheemtourreg}
\end{figure}

Now that we have covered the classification and regression cases we will discuss the preprocessing, package infrastructure, and interactive features before proceeding to case studies with data from the wild.

The dynamic interaction with the global view is critical to the selection of the primary and comparison observations. Linked brushing also for a rectangular selection of observations. These points are highlighted within the other space of the global view \emph{and} crucially available for downstream consumption, for instance, highlighting in the tour. The distinction between the former and latter is subtle, but important for developers to grasp. In the former, self-contained HTML widgets can highlight directly within javascript without evaluation downstream in other reactive functions. In the latter, the use of \texttt{plotly::event\_data()} returns the identity of the selected observations. This will require a reactive flush, but after that, the sky is the ceiling.

@@

The benefit of having dynamic interaction with data is predicated on a reasonably small render time. It is important to preprocess as much work as possible so that application resources can be used efficiently. The work remaining at runtime should be solely responding to inputs and the rendering of figures and tables. Below we discuss the steps and details of the reprocessing.

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable, an optional aesthetic variable (color/shape)  can be mapped typically a categorical variable.
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} (Liaw and Wiener 2002) to mitigate the runtime of our local explanation which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP applicable to tree-based models. This is done with the package \textbf{treeshap} (Kominsarczyk et al. 2021), hosted on GitHub only]. The global view shows all observations in attribution space requiring that we must extract the variable weightings from \emph{all} observations rather than just one.
    <!-- \item \textbf{Global view:} The data- and attribution-spaces are approximated as their the first two principal components. -->
\end{itemize}

The time to preprocess the data will vary significantly with the choice of model and local explanation. For reference, the FIFA data, 5000 observations of 9 explanatory variables, took 2.9 seconds to fit a random forest models of modest hyperparmeters. Extracting the tree SHAP values of each observation took 254 seconds combined. PCA and statistics of the variables and attribution took 0.6 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice it to say that the bulk of the time will be spent on the local attribution. This makes tree SHAP, with its reduced comupational complexity a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

The above-described method and application are implemented as an open-source \textbf{R} package, \textbf{cheem} \emph{TODO:XXX include cran URL after published}. Preprocessing was facilitated with models created via \textbf{randomForest} {[}liaw\_classification\_2002{]}, and explanations calculated with \textbf{treeshap} (Kominsarczyk et al. 2021). The application is made with \textbf{shiny} (Chang et al. 2021). The tour visual is an extension of \textbf{spinifex} (Spyrison and Cook 2020). Both views are created first with first with \textbf{ggplot2} (Wickham 2016) and then rendered as interactive HTML widgets with \textbf{plotly} (Sievert 2020). \textbf{DALEX} (Biecek 2018) and the free ebook, \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021) was a huge boon to understanding local explanations and how to apply them.

@@

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# Download the package:}
\FunctionTok{install.packages}\NormalTok{(}\StringTok{"cheem"}\NormalTok{, }\AttributeTok{dependencies =} \ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{library}\NormalTok{(}\StringTok{"cheem"}\NormalTok{)}
\DocumentationTok{\#\# Try the app:}
\FunctionTok{run\_app}\NormalTok{()}
\DocumentationTok{\#\# Bring your own data: follow the examples in cheem\_ls()}
\NormalTok{?cheem\_ls}
\end{Highlighting}
\end{Shaded}

@@

To illustrate the use of the cheem method we apply it to modern datasets, two classification examples and then two of regression.

@@

\caption{Chocolates data  type classification (milk or dark). }\label{fig:casechocolates}
\end{figure}

>>>>

  article]{article}

@@

\begin{document}
\maketitle

@@

Mathematically rigorous approaches to predictive modeling are attributed to the least-squares method, over two centuries ago by Legendre and Gauss in 1805 and 1809, respectively. In 1886 Francis Galton coined the term \emph{regression} to refer to continuous, quantitative predictions. While \emph{classification} refers to discrete predictions as introduced by Fisher in 1936.

Breiman and Shmueli (Breiman 2001; Shmueli 2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose such as hypothesis testing, while \emph{predictive} modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, interpretability is key for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use has important implications for model selection and development.

Black-box models are becoming increasingly common, but not without their share of controversy and issues (O'neil 2016; Kodiyan 2019). Black-box models have been known to reflect common biases, including sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018). Such issues occur when biases existent in the training data, the model picks up on this influence on the response variable, which is then built into the model. Another issue is data drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions (Lazer et al. 2014; Salzberg 2014). Such cases highlight the need to make models fair, accountable, ethical, and transparent, which has led to the movement of XAI (Adadi and Berrada 2018; Arrieta et al. 2020).

One branch of XAI is local explanations, which take a variable attribution approach to bring transparency to a model. Local explanations attempt to approximate linear variable importance at the location of one observation. There are many such local explanations.

To illustrate our work, we apply the model-agnostic explanation SHAP (Strumbelj and Kononenko 2010; Štrumbelj and Kononenko 2014). The exact details of SHAP are tangent to the ideas of this work, but suffice it to say that SHAP approximates variable importance by taking the median importance over permutations of the explanatory variables. To be exact, we apply a variant that enjoys a lower computational complexity, known as tree SHAP (S. M. Lundberg, Erion, and Lee 2018).

In multivariate data visualization, a \emph{tour} (Asimov 1985; Buja and Asimov 1986; S. Lee et al. 2021) is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

There are various types of tours, which are distinguished by the generation of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020) this path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path (Y. D. Lee et al. 2013; Silva, Cook, and Lee 2021).

The approach proposed below is to use the radial, manual tour to interrogate a local explanation. After the identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package \textbf{cheem} with an interactive application to facilitate analysis. We give case studies of toy and modern datasets for both classification and regression tasks.

The change in the projection basis might feel similar to counterfactual, what-if analysis, such as \emph{ceteris paribus} (Biecek 2020). Latin for ``other things held constant'' or ``all else unchanged'' is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual, observed case by that varies contributions of the basis, essentially the orientation of the data object. Another difference is that the basis must maintain orthonormality. That is to say, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction.

The remainder of this paper is organized as follows. The next section, \protect\hyperlink{sec:explanations}{Local explanation statistics}, covers the background of the local explanation, SHAP, and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into these animations of continuous linear projections. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the layout of the application, how it facilitates analysis, preprocessing, and package infrastructure. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:discussion}{Discussion} of the insights we draw from classification and regression tasks.

@@

Consider a highly non-linear model. At face value, it is hard to say which variable(s) are sensitive to the crossing of a classification boundary or identifies which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

Figure 6 of Arrieta et al. (2020) gives comprehensive summarization of the taxonomy and literature of explanation techniques. This includes a large number of model-specific explanations such as deepLIFT, (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017) a popular recursive method for estimating importance in neural networks. There are a fewer number of model-agnostic explanations, of which LIME, (Ribeiro, Singh, and Guestrin 2016) SHAP, (S. Lundberg and Lee 2017) and their variants are popular.

These instance-level explanations are used in a variety of ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were important (Simonyan, Vedaldi, and Zisserman 2014). For instance, the presence of snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain variable additive contributions from model intercept to the observation's prediction (Ribeiro, Singh, and Guestrin 2016).

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation by taking the median importance of a subset of permutations in the explanatory variables. This idea stems from the field of game theory, where Shapley (1953) devised a method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score.

To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season (Leone 2020). We have 5000 observations of nine skill measures (after aggregating highly correlated variables). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more important to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry in the variable ordering, so we opt to fix the order to that of panel b), namely, by decreasing the sum of the SHAP values.

@@

\caption{Illustration of the distribution of SHAP attributions and a breakdown plot. From FIFA 2020 data, a random forest model regresses wages from nine skill attributes for a star offensive and defensive player. The players have very different wages, but b) shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The medians of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the player's position. c) Breakdown plots of the observations the explanation used to additively explain the difference between the model intercept and the observation prediction.}\label{fig:shapdistrbd}
\end{figure}

In summary, this highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, we showed how two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In the following section, we will be using normalized explanations as the starting projection basis to interrogate the explanation further.

@@

Below we illustrate the two primary displays of the application: the global view and the tour view. Then we cover what we take away from the classification and regression tasks. Lastly, we discuss the preprocessing that before display.

The global view provides an essential context of all observations and allows exploration of the separability of the data- and attribution-spaces. While ultimately, its purpose is to facilitate the selection of a primary point and comparison point.

An approximation of these spaces is given as the first two principal components of their respective spaces. This is shown side-by-side next to model information, the prediction, and the observed response variable. The orientation and magnitude of the variables are inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it is a reasonable summarization real task at hand, the selection of observation, and nearby comparison.

It is insightful to explore these two approximations against a visual of the model; prediction by observation is also displayed. Linked brushing and preserved aesthetic features such as circling misclassified observations help link information from the different spaces together.

\hypertarget{radial-cheem-tour}{%
\subsection{Radial cheem tour}\label{radial-cheem-tour}}

The global view facilitated the selection of a primary and optional comparison observation. The variable-level attribution of primary observation is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear variables that best explain the difference between the model intercept and an observations prediction, not the local shape of the model surface.

The initial frame is the normalized SHAP values of the primary observation. The current projection basis is depicted as the width of a bar, the variable's contribution to the horizontal axis. The normalized values of all observations are shown as vertical parallel coordinate plots.

The radial tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing so tests an individual variable's sensitivity to the structure identified by the local explanation. The default variable selected has the largest discrepancy between the primary and comparison observations attribution. In the following sections, we elaborate on the takeaways we draw from applying this approach in classification and regression tasks, respectively.

Now that we have introduced the global view and corresponding cheem radial tour let's discuss the differences between the classification and regression cases.

@@

What information do we glean from using this method on a classification task? Typically we select a misclassified observation compared to a correctly classified point that is nearby in data space. We start by seeing the data projected through the linear attribution, the combination that best justifies that prediction. By default, the manual tour varies the contribution of the variable with the largest difference between the primary and comparison observation. That is, we can test the sensitivity of each variable to structure identified by the local explanation; we are exploring the support of the explanation, evaluating the support or robustness of the prediction.

{\centering \includegraphics[width=1\linewidth]{./figures/app_classification}

\caption{Display illustrating the classification case. Both views are colored on predicted class and while red circles identify misclassified observations. The radial tour is a 1D projection starting at the normalized tree SHAP values of the primary point. The first frame is the linear-variable importances that best describe the difference from model intercept to this observation's prediction. We probe the support of the variable contributions by selecting a variable to vary the contribution.}\label{fig:classificationcase}
\end{figure}

@@

The regression case, we opt to color the global view on a statistic, the observation's residual, the log Mahanabis distance in data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the attribution projection, then varying as the radial tour changes basis. While the vertical position is fixed to the observed response variable and residuals for the middle and right panels, respectively. This changes the display from 1D density to a scatterplot. The basis is still one component, the horizontal position, independent with the vertical position.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/app_regression}

@@

\caption{Display of the regression case. The global view can be colored on the correlation of the attribution projection and observed response. In the tour, the horizontal values are the same as the classification case. The vertical position is fixed to the observed response and residuals in the middle and right facets, respectively.}\label{fig:regressioncase}
\end{figure}

The application has several reactive selections that affect the data, coloring, and inputs for the tour to improve flexibility by extending the use cases. It also has interactive features in the global view that aid the analysis.

A tooltip displays while the cursor hovers over a point displays the observation number/name and classification information if appropriate. Linked brushing allows for the selection of points (with left click and drag) where those points will be highlighted in both plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and ultimately the selection of the primary and comparison observations.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/app_interactions}

@@

\caption{Illustration of data explorations interactions in the global view. This view has linked brushing of the points where observations selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over a point}\label{fig:interactions}
\end{figure}

@@

The benefit of having dynamic interaction with data is predicated on a reasonably short render time. It is important to preprocess expensive operations so the application resources can be used efficiently. The work remaining at runtime is solely responding to inputs and rendering of visuals and tables. Below we discuss the steps and details of the reprocessing.

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to color and shape of obervations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)] for compatability with the local explanation which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, that is applicable to tree-based models. This is done with the package \textbf{treeshap} [Kominsarczyk et al. (2021), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 observations of nine explanatory variables, took 2.9 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 254 seconds combined. PCA and statistics of the variables and attributions took 0.6 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice it to say that the bulk of the time will be spent on the local attribution. Increased model complexity or data dimensionality will quickly become an obstacle. This makes tree SHAP, with its reduced computational complexity, a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

The above-described method and application are implemented as an open-source \textbf{R} package, \textbf{cheem} \emph{TODO:XXX include cran URL after published}. Preprocessing was facilitated with models created via \textbf{randomForest} {[}liaw\_classification\_2002{]}, and explanations calculated with \textbf{treeshap} (Kominsarczyk et al. 2021). The application was made with \textbf{shiny} (Chang et al. 2021). The tour visual is an extension of \textbf{spinifex} (Spyrison and Cook 2020). Both views are created first with first with \textbf{ggplot2} (Wickham 2016) and then rendered as interactive HTML widgets with \textbf{plotly} (Sievert 2020). \textbf{DALEX} (Biecek 2018) and the free ebook, \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021) was a huge boon to understanding local explanations and how to apply them.

@@

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# Download the package}
\FunctionTok{install.packages}\NormalTok{(}\StringTok{"cheem"}\NormalTok{, }\AttributeTok{dependencies =} \ConstantTok{TRUE}\NormalTok{)}
\DocumentationTok{\#\# Restart the R session to pick up the file structure may be needed}
\FunctionTok{restartSession}\NormalTok{()}
\DocumentationTok{\#\# Load cheem into session}
\FunctionTok{library}\NormalTok{(}\StringTok{"cheem"}\NormalTok{)}
\DocumentationTok{\#\# Try the app}
\FunctionTok{run\_app}\NormalTok{()}
\DocumentationTok{\#\# Bring your own data: follow the examples in cheem\_ls()}
\NormalTok{?cheem\_ls}
\end{Highlighting}
\end{Shaded}

@@

To illustrate the use of the cheem method, we apply it to modern datasets, two classification examples and then two of regression.

@@

\caption{Chocolates data type classification (milk or dark).}\label{fig:casechocolates}
\end{figure}
